Left-linear rules in adhesive categories

### 2024-02-12 icalp/fusione.tex
miniri

\begin{abstract}
  When can two sequences of steps executed by a computational device
  be considered equivalent?  This is a relevant question if one
  considers concurrent and distributed systems, since the independence
  of two steps means that they could be executed in any order and
  potentially in parallel.
  % 
  We investigate the issue in the context of adhesive categories, a
  very general instance of the DPO approach to rewriting. More
  precisely, we focus on left-linear rules, which not only consume,
  preserve and generate entities, as linear rules do, but also have
  the capability of ``merging'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since the canonical notion of independence used for linear rewriting
  systems does not work out of the box.
  %
  Moreover, a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds, and this requires a substantial revision of the
  theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

Among rewriting-based formalisms, those manipulating graph-like structures
proved useful in many settings for representing the dynamics of distributed systems: the 
graph serves to represent the entities and their relations within states, while 
rewriting steps, modifying the graph, model computation steps that result 
in changes to the system state. The DPO (double-pushout) approach~\cite{CMREHL:AAGT} 
is nowadays considered the standard one for these structures, thanks to its locality 
(rule application has a local effect on a state, differently e.g. from the SqPO 
approach~\cite{CorradiniHHK06}) and most importantly to its flexibility: the rules allow to 
specify that,  in a rewriting step, some entities in the current state are removed, 
some others are needed for the rewriting step to occur but remain unaltered and 
some new entities can be generated. Concerning concurrency, the chosen notion 
is shift equivalence, and independence is formally captured by the so-called 
Church-Rosser theorem of DPO rewriting~\cite[Section~xxx]{CMREHL:AAGT}.

A noteworthy advantage is that DPO rewriting can be formulated over
adhesive categories and their variants~\cite{lack2005adhesive,ehrig2006weak}. 
%
Operating within the framework of adhesive categories allows one to devise 
the foundations of a rewriting theory that can be subsequently instantiated 
to the context of interest, thus avoiding the need of repeatedly proving
similar ad-hoc results in for each specific setting. In fact, the framework is
so general that it allowed to recast in the DPO approach many additional formalisms,
ranging from toposes~\cite{johnstone2007quasitoposes} to string diagrams~\cite{bonchi2022string}.

So far, most of the theoretical results focussed on what are called \emph{linear rules}: 
intuitively, a step is required to only consume, preserve and generate entities. 
However, in some situations it is also necessary to ``merge'' parts of the state. 
This happens already in the context of string diagrams mentioned above, but also 
in graphical implementations of nominal calculi where, as a result of name passing, 
the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the visual modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expression, the idea
here -- conceptually and, as we will see, also mathematically similar --
is to maintain an equivalence\todo{P: more precisely a congruence}
over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
  vedere se c'\`e qualcosa d'altro}

@@

Technically, to acquire the expressiveness needed for expressing
fusions, i.e., the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, to the best of our knowledge the
phenomena arising when rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories 
%(which capture most ``applications'') 
have not been systematically explored. ADD REFS

@@

  We will say that two derivations $\der{D}, \der{E}:G\Mapsto H$ are \emph{switch
    equivalent} if there is a switching sequence, possibly empty, $\{\der{D}_{k}\}_{i=0}^m$ such that
      \[\der{D}\equiv_a \der{D}_0 \shift{\transp{i_1}} \der{D}_1 \shift{\transp{i_2}} \ldots \shift{\transp{i_m}}
    \der{D}_m\equiv _a \der{E}\]
\end{definition}

@@

\begin{restatable}{lemma}{lemIde}\label{lem:ide}
	Let $\der{D}$ and $\der{D}'$ be two derivations in a linear DPO-rewriting system $(\X, \R)$ with the same length $l$. If $\id{[0,l-1]}:[0,l-1]\to [0,l-1]$ is a consistent permutation, then $\der{D}$ and $\der{D}'$ are abstraction equivalent.  
\end{restatable}
[Proof in \Cref{lemide-proof}]

@@

\begin{corollary}
	\label{cor:ele}
	Let $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ be a derivation between $G$
	and $H$. Let $j$ and $k$ be two indexes less or equal than $n+1$ and
	suppose that $j< k$.  Consider two arrows $a\colon T\to G_j$,
	$b\colon T\to G_k$. If $\iota_{G_j}\circ a = \iota_{G_k}\circ b$,
	then there exist a unique arrow $c\colon T\to D_j $ such
	that \[f_j\circ c = a\qquad \iota_{D_j}\circ c =\iota_{G_k}\circ b\]
\end{corollary}

\lemIde*
\label{lemide-proof}
\begin{proof}
	If $\der{D}$ and $\der{D}'$ are empty then the thesis follows at once since an abstraction equivalence is just an isomorphism. Let now $\der{D}$ and $\der{D}'$ be, respectively $\{\dder{D}_{i}\}_{i=0}^n$ and $\{\dder{D}'_i\}_{i=0}^n$. We construct by induction on $i\in [0,n]$ an abstraction equivalence such that, for every $i$, $\iota'_{G'_{i+1}}\circ \phi_{G_{i+1}}=\xi_{\id{[0,n]}}\circ \iota_{G_{i+1}}$.

@@

	To conclude, it is now enough to notice that
	\begin{align*}
		&\iota'_{G'_{i+1}}\circ \phi_{G_{i+1}}\circ g_i=\iota'_{G'_{i+1}}\circ g'_i\circ \phi_{D_{i}}=\iota'_{D_{i}}\circ \phi_{D_{i}}=\iota'_{G_i}\circ f'_i\circ \phi_{D_{i}} =\iota'_{G'_{i}}\circ \phi_{G_{i}}\circ f_i\\=&\xi_{\id{[0,n]}}\circ \iota_{G_{i}}\circ f_i=\xi_{\id{[0,n]}}\circ \iota_{D_{i}}=\xi_{\id{[0,n]}}\circ \iota_{G_{i+1}}\circ g_i
	\end{align*}
	Which, 	since $G_{i}$ is obtained pushing $r_{i-1}$ along $k_{i-1}$ gives us the thesis 
\end{proof}

@@

\begin{lemma}
	Let $\der{D}$ and $\der{E}$ be two derivations in a linear rewriting system $(\X, \R)$. Suppose that a consistente permutation $\sigma:[0, \lgh(\der{D})-1]\to [0, \lgh(\der{D})-1]$ between them is given. If $i\in [0, \lgh(\der{D})-1]$ is an index such that $\transp{i}$ is an inversion for $\sigma$, then the direct derivations $\dder{D}_i$ and $\dder{D}_{i+1}$ are sequentially independent.
\end{lemma}
\begin{proof}
	contenuto...
\end{proof}

\begin{remark}\label{rem:salvavita}
	Let 
\end{remark}
\thmcoswitch*
\label{thmswitch-proof}
\begin{proof}$(1\Rightarrow 2)$ This follows from \Cref{rem:abs,prop:coswitch}.

>>>>

\begin{abstract}
  When two sequences of steps executed by a computational device can
  be considered equivalent?  This is a relevant question if one
  considers concurrent and distributed systems, since the independence
  of two steps means that they could be executed in any order, and
  potentially in parallel.
  % 
  We investigate the issue in the context of adhesive categories, a
  very general instance of the DPO approach to rewriting. More
  precisely, we focus on left-linear rules, which not only consume,
  preserve and generate entities, as linear rules do, but may
  ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since the canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to well-behaved classes of left-linear rules.
  %
  In fact, the standard characterisation of independence does not hold anymore,
  and this requires a substantial revision of the 
  theory for the linear case, since a number of basic properties,
  essential in the development of a theory of concurrency, were
  derived from such a description.
  %
  The paper introduce a novel characterisation of independence, for which we prove a Church-Rosser property,
  properly extending the class of formalisms 
  that can be modelled in the adhesive framework. 
\end{abstract}

@@

Among rewriting-based formalisms, those manipulating graph-like structures
proved useful in many settings for representing the dynamics of systems: the 
graph serves to represent the entities and their relations within states, while 
rewriting steps, modifying the graph, model computation steps that result 
in changes to the system state. The DPO (double-pushout) approach~\cite{CMREHL:AAGT} 
is nowadays considered the standard one for these structures, thanks to its locality 
(rule application has a local effect on a state, differently e.g. from the SqPO 
approach~\cite{CorradiniHHK06}) and most importantly to its flexibility: the rules allow to 
specify that,  in a rewriting step, some entities in the current state are removed, 
some others are needed for the rewriting step to occur but remain unaltered and 
some new entities can be generated. Concerning concurrency, the chosen notion 
is shift equivalence, and independence is formally captured by the so-called 
Church-Rosser theorem of DPO rewriting~\cite[Section~xxx]{CMREHL:AAGT}.

A noteworthy advantage is that DPO rewriting can be formulated over
adhesive categories and their variants~\cite{lack2005adhesive,ehrig2006weak}. 
%
Operating within the framework of adhesive categories allows one to devise 
the foundations of a rewriting theory that can be subsequently instantiated 
to the context of interest, thus avoiding the need of repeatedly proving
similar ad-hoc results in for each specific setting. In fact, the framework is
so general that it allowed to recast in the DPO approach many additional formalisms,
ranging from xxx to string diagrams~\cite{bonchi2022string}.

So far, most of the theoretical results focussed on what are called \emph{linear rules}: 
intuitively, a step is required only to consume, preserve and generate entities. 
However, in some situations it is also necessary to ``merge'' parts of the state. 
This happens already in the context of string diagram mentioned above, but also 
in graphical implementations of nominal calculi where, as a result of name passing, 
the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the visual modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expression, the idea
here - conceptually and, as we will see, also mathematically similar -
is to maintain an equivalence\todo{P: more precisely a congruence}
over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
  vedere se c'\`e qualcosa d'altro}

@@

Technically, to acquire the expressiveness needed for expressing
fusions, i.e., the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, to the best of our knowledge the
phenomena arising when rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories (which capture most
``applications'') have not been systematically explored. ADD REFS

@@

  We will say that two derivations $\der{D}, \der{E}:G\Mapsto H$ are \emph{switch
    equivalent} if there is a switching sequence, possibly empty, $\{\der{D}_{k}\}_{k=0}^m$ such that
      \[\der{D}\equiv_a \der{D}_0 \shift{\transp{i_1}} \der{D}_1 \shift{\transp{i_2}} \ldots \shift{\transp{i_m}}
    \der{D}_m\equiv _a \der{E}\]
\end{definition}

@@

\begin{restatable}{proposition}{lemIde}\label{lem:ide}
	Let $\der{D}$ and $\der{D}'$ be two derivations in a linear DPO-rewriting system $(\X, \R)$ with the same length $l$. If $\id{[0,l-1]}:[0,l-1]\to [0,l-1]$ is a consistent permutation, then $\der{D}$ and $\der{D}'$ are abstraction equivalent.  
\end{restatable}
[Proof in \Cref{lemide-proof}]

@@

\begin{corollary}
	\label{cor:ele}
	Let $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ be a derivation between $G$
	and $H$. Let $j$ and $k$ be two indexes less or equal than $n+1$ and
	suppose that $j< k$.  Consider two arrows $a\colon T\to G_j$,
	$b\colon T\to G_k$. If $\iota_{G_j}\circ a = \iota_{G_k}\circ b$,
	then there exist a unique arrow $c\colon T\to D_j $ such
	that \[f_j\circ c = a\qquad \iota_{D_j}\circ c =\iota_{G_k}\circ b\]
\end{corollary}
\begin{lemma}\label{lem:idevero}
	Let $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ and $\der{D}'=\{\dder{D}'_i\}_{i=0}^n$ be two non-empty derivations in a linear DPO-rewriting system $(\X, \R)$ . Let  $\sigma:[0,n-1]\to [0,n-1]$  is a consistent permutation and suppose that the set 
	\[E_{\sigma}=\{i\in [0,n-1]\mid \sigma(i)=i\}\]
	is non empty with maximum $k$. Then $\{\dder{D}_i\}_{i=0}^k$ and $\{\der{D}'_i\}_{i=0}^k$ are abstraction equivalent.  
\end{lemma}

\begin{proof}We construct are going to construct by induction on $i\in [0,n]$ an abstraction equivalence such that, for every $i$, $\iota'_{G'_{i+1}}\circ \phi_{G_{i+1}}=\xi_{\id{[0,n]}}\circ \iota_{G_{i+1}}$.

	\smallskip \noindent \parbox{8.2cm}{$i=0$. We start noticing that, by consistency
		\[\iota'_{G_0}\circ m'_0=\xi_\sigma \circ \iota_{G_0}\circ m_0=\iota'_{G_0}\circ m_0\]
		By linearity and \Cref{lem:colim}, $\iota'_{G_0}$ is a mono, thus $m_0=m'_0$. By  \Cref{prop:unique} we deduce the existence of the dotted isomorphisms in the diagram aside.} \parbox{4cm}{\vspace{0ex}$\xymatrix@C=40pt{G_0&D'_0 \ar@{>->}[r]^{g'_0}
			\ar@{>->}[l]_{f'_0}&G'_{1}\\ L_0 \ar[u]^{m'_0} \ar[d]_{m_0}& K_0
			\ar[u]^{k'_0} \ar[d]_{k_0} \ar@{>->}[r]^{r_0} \ar@{>->}[l]_{l_0}
			&R_0\ar[u]^{h'_0} \ar[d]_{h_0}\\G_0
			\ar@/_.45cm/[uu]_(.35){\id{G_0}}|\hole& D_0 \ar@{>->}[l]^{f_0}  \ar@{.>}@/_.45cm/[uu]_(.35){\phi_{D_0}}|\hole
			\ar@{>->}[r]_{g_0}& G_{1}\ar@{.>}@/_.45cm/[uu]_{\phi_{G_1}}}$}

	Now, if we compute we have
	\begin{align*}
		&\iota'_{G'_1}\circ \phi_{G_1}\circ g_0=\iota'_{G'_1}\circ g'_0\circ \phi_{D_0}=\iota'_{D_0}\circ \phi_{D_0}=\iota'_{G_0}\circ f'_0\circ \phi_{D_0}\\=&\iota'_{G_0}\circ f_0=\xi_{\sigma} \circ \iota_{G_0}\circ  f_0=\xi_{\sigma} \circ \iota_{D_0}=\xi_{\sigma} \circ \iota_{G_1}\circ g_0 
	\end{align*}
	This now entails that $\iota'_{G'_1}\circ \phi_{G_1}=\xi_{\id{[0,n]}}\circ \iota_{G_1}$.

	\smallskip  \noindent \parbox{7.5cm}{$i>0$. By the inductive hypothesis there exists an isomorphism $\phi_{G_{i}}\colon G_{i}\to G'_i$ such that $\phi_{G_{i}}\circ h_{i-1}=h'_{i-1}$. Now, as before we have
		\[\iota'_{G'_{i}}\circ m'_{i}=\xi_{\sigma}\circ \iota_{G_i}\circ m_{i}=\iota'_{G'_{i}}\circ \phi_{G_{i}} \circ m_i \]
		By linearity $\iota'_{G'_{i}}$ is mono, thus $m'_i=\phi_{G_{i}} \circ m_i$ and \Cref{prop:unique} yields again the dotted part of the diagram aside.}
	\parbox{4cm}{\vspace{-2em}$\xymatrix@C=40pt{G_i&D'_i \ar@{>->}[r]^{g'_i}
			\ar@{>->}[l]_{f'_i}&G'_{i+1}\\ L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i
			\ar[u]^{k'_i} \ar[d]_{k_i} \ar@{>->}[r]^{r_i} \ar@{>->}[l]_{l_i}
			&R_i\ar[u]^{h'_i} \ar[d]_{h_i}\\G_i
			\ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole& D_i \ar@{>->}[l]^{f_i}  \ar@{.>}@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole
			\ar@{>->}[r]_{g_i}& G_{i+1}\ar@{.>}@/_.45cm/[uu]_{\phi_{G_{i+1}}}}$}

	To conclude, it is now enough to notice that
	\begin{align*}
		&\iota'_{G'_{i+1}}\circ \phi_{G_{i+1}}\circ g_i=\iota'_{G'_{i+1}}\circ g'_i\circ \phi_{D_{i}}=\iota'_{D_{i}}\circ \phi_{D_{i}}=\iota'_{G_i}\circ f'_i\circ \phi_{D_{i}} \\=&\iota'_{G'_{i}}\circ \phi_{G_{i}}\circ f_i=\xi_{\sigma}\circ \iota_{G_{i}}\circ f_i=\xi_{\sigma}\circ \iota_{D_{i}}=\xi_{\sigma}\circ \iota_{G_{i+1}}\circ g_i
	\end{align*}
	Which, 	since $G_{i+1}$ is obtained pushing $r_{i}$ along $k_{i}$ gives us the thesis 
\end{proof}

From \Cref{lem:idevero} we can deduce at once the following.

\lemIde*
\label{lemide-proof}
\begin{proof}
	If $\der{D}$ and $\der{D}'$ are empty then the thesis follows at once since an abstraction equivalence is just an isomorphism. Otherwise, if $\der{D}$ and $\der{D}'$ are non-empty the thesis follows from \Cref{lem:idevero}.

	\iffalse 
	We construct by induction on $i\in [0,n]$ an abstraction equivalence such that, for every $i$, $\iota'_{G'_{i+1}}\circ \phi_{G_{i+1}}=\xi_{\id{[0,n]}}\circ \iota_{G_{i+1}}$.

@@

	To conclude, it is now enough to notice that
	\begin{align*}
		&\iota'_{G'_{i+1}}\circ \phi_{G_{i+1}}\circ g_i=\iota'_{G'_{i+1}}\circ g'_i\circ \phi_{D_{i}}=\iota'_{D_{i}}\circ \phi_{D_{i}}=\iota'_{G_i}\circ f'_i\circ \phi_{D_{i}} =\iota'_{G'_{i}}\circ \phi_{G_{i}}\circ f_i\\=&\xi_{\id{[0,n]}}\circ \iota_{G_{i}}\circ f_i=\xi_{\id{[0,n]}}\circ \iota_{D_{i}}=\xi_{\id{[0,n]}}\circ \iota_{G_{i+1}}\circ g_i
	\end{align*}
	Which, 	since $G_{i}$ is obtained pushing $r_{i-1}$ along $k_{i-1}$ gives us the thesis 
\fi 
\end{proof}

@@

\begin{lemma}
	Let $\der{D}$ and $\der{D}$ be two derivations in a linear rewriting system $(\X, \R)$. Suppose that a consistent permutation $\sigma:[0, \lgh(\der{D})-1]\to [0, \lgh(\der{D})-1]$ between them is given. If $i\in [0, \lgh(\der{D})-1]$ is the smallest index such that $\transp{i}$ is an inversion for $\sigma$, then the direct derivations $\dder{D}_i$ and $\dder{D}_{i+1}$ are sequentially independent.
\end{lemma}
\begin{proof}
	Let us start with the diagram below
	  \[\xymatrix@C=15pt{L_i \ar[d]_{m_i}&& K_i
		\ar[d]_{k_i}\ar@{>->}[ll]_{l_i} \ar@{>->}[r]^{r_i} & R_i
		\ar@{.>}@/^.35cm/[drrr]|(.3)\hole_(.4){j_0}
		\ar[dr]|(.3)\hole_{h_i} && L_{i+1} \ar@{.>}@/_.4cm/[dlll]^(.4){j_1}
		\ar[dl]|(.3)\hole^{m_{i+1}}& K_{i+1} \ar[d]^{k_{i+1}}\ar@{>->}[l]_{l_{i+1}}
		\ar@{>->}[rr]^{r_{i+1}} && R_{i+1} \ar[d]^{h_{i+1}}\\G_i && \ar@{>->}[ll]^{f_i}
		D_i \ar@{>->}[rr]_{g_i}&& G_{i+1} && \ar@{>->}[ll]^{f_{i+1}} D_{i+1}\ar@{>->}[rr]_{g_{i+1}}&& G_{i+2}}\]
	\parbox{9cm}{Notice that $\sigma$ restrict to the identity $[0, i-1]\to [0,i-1]$, which remains consistent. In particular, there exists an isomorphism $\phi_{i}\colon G_i\to G'_i$ such that the square on the right commutes.}
	\parbox{3cm}{\vspace{-.5cm}$\xymatrix{G_i \ar[d]_{\iota_{G_i}}\ar@/^.4cm/[rr]^{\phi_{G_i}}& L_0  \ar[l]^{m_i}\ar[r]_{m'_{i}} & G'_{i} \ar[d]^{\iota'_{G'_i}}\\ \tpro{D} \ar[rr]_{\xi_\sigma} & &\lpro\der{D}'\rpro  }$}

	By consistency we have the following equations:
	\[\begin{split}
		\iota_{G'_\sigma(i+1)}\circ m'_{\sigma(i+1)} =\xi_{\sigma}\circ \iota_{\sigma(i)}
	\end{split}\]
\end{proof}

\begin{remark}\label{rem:salvavita}
	Before proceeding further, it is worth to point out the following fact about inversions.
\end{remark}
\thmcoswitch*
\label{thmswitch-proof}
\begin{proof}$(1\Rightarrow 2)$ This follows from \Cref{rem:abs,prop:coswitch}.
